---
title: "EEB331 Project 13 FINAL Code"
author: "Leila Granier and Yara Almoussa"
date: "2026-04-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(stringr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(MuMIn)
library(kableExtra)
library(cowplot)
```

Aesthetics
```{r}
eeb_theme <- function(base_size = 9, base_family = "Arial") {
  theme_classic(base_size = base_size, base_family = base_family) +
    theme(
      # Set all text to Arial
      text = element_text(family = "Arial"),
      # Remove grid lines and background
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "transparent", color = NA),
      plot.background = element_rect(fill = "transparent", color = NA),
      # Plot title
      plot.title = element_text(size = base_size, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = base_size, hjust = 0.5, margin = margin(b = 10)),
      # Axes
      axis.line = element_line(color = "black"),
      axis.text = element_text(size = 9, color = "black"),
      axis.title = element_text(size = 9),
      # Legend
      legend.background = element_rect(fill = "transparent", color = NA),
      legend.key = element_blank(),
      legend.text = element_text(size = 9),
      legend.title = element_text(size = 9),
    ) 
}
```

Fraction of medium-long ROH Calculation (from Precept)
```{r}
# Constant
genome_size <- 2319465413 # in nucleotide base pairs (calculated)

# Read in the ROH output from bcftools
lines <- readLines("Marmots_maf3_geno90_mind20_ROH.txt")
lines <- lines[!grepl("^#", lines)]
rg_lines <- lines[grepl("^RG", lines)]

# Extract ROH segments based on file structure
roh_segments <- read.table(
  text = rg_lines,
  header = FALSE,
  stringsAsFactors = FALSE)

# Add back in column names based on read group (RG) headers
colnames(roh_segments) <- c(
  "type",
  "sample",
  "chr",
  "start",
  "end",
  "length_bp",
  "n_markers",
  "quality")

# Calculate some basic metrics
long_roh_summary <- roh_segments %>%
  mutate(length_bp = as.numeric(end) - as.numeric(start)) %>%
  # Filtering for ROHs longer than 1 Mb = 1,000,000 bp (medium to long ROHs)
  filter(length_bp > 1000000) %>% 
  group_by(sample) %>%
  summarise(
    NSEG = n(),
    total_bp = sum(length_bp),
    mean_bp = mean(length_bp),
    mean_length = total_bp / NSEG)

# Calculate FROH per individual
# FROH = the fraction of an individual's genome that is identical by descent (IBD)
long_roh_summary <- long_roh_summary %>%
  filter(str_starts(sample, "UID")) %>% 
  mutate(FROH = total_bp / genome_size) %>%  
  mutate(uid = sub("UID_(\\d+_\\d+)_sorted.*", "\\1", sample))
head(long_roh_summary)
```

Combining marmot mass data, metadata, and FROH data
```{r}
# Import data files
marm_mass <- read.csv("marmot_mass_metadata.csv")
metadata <- read.csv("RADseq_pedigree_metadata.csv")

# Merge data sets to make final data frame
mass_FlongROH <- merge(metadata, merge(long_roh_summary, marm_mass, by = "uid"), by = "uid") %>% 
  mutate(uid = as.factor(uid),
         sex = as.factor(sex),
         age = as.factor(age),
         yrborn = as.factor(yrborn)) %>% 
  select(uid, sex, yrborn, age, FROH, massaug, massjun,  nb_mass, total_bp, 
         mean_bp, mean_length, NSEG, col_area) %>% 
  mutate(massgain = massaug-massjun,
         mean_length_kb = mean_length/1000)
mass_FlongROH
```

FROH vs seasonal mass gain
AIC analysis to identify best model
```{r}
# Linear model of mass gain with just FROH as fixed effect
base_mod <- lm(massgain ~ FROH, data = mass_FlongROH)

# Linear model of mass gain with FROH and sex as fixed effects
mod_sex <- lm(massgain ~ FROH + sex, data = mass_FlongROH)

# Linear mixed effects model of mass gain with FROH as fixed effect and year born as random effect
mod_yrborn <- lmer(massgain ~ FROH + (1|yrborn), 
                    data = mass_FlongROH, REML = FALSE)

# Linear mixed effects model of mass gain with FROH and sex as fixed effects and year born as random effect
mod_sexyrborn <- lmer(massgain ~ FROH + sex + (1|yrborn), 
                       data = mass_FlongROH, REML = FALSE)


# Conduct AIC analyses
ms_df <- as.data.frame(model.sel(base_mod, mod_sex, mod_yrborn, mod_sexyrborn))

# Keep only useful columns
ms_df <- ms_df[, c("df", "logLik", "AICc", "delta", "weight")]

ms_df$AICc <- round(ms_df$AICc, 1)
ms_df$delta <- round(ms_df$delta, 3)

# Rename columns and models
colnames(ms_df) <- c("K", "logLik", "AICc", "Delta AICc", "Weight")

ms_df$Model <- c("Base model + birth year (random)",
  "Base model + sex (fixed) + birth year (random)",
  "Base model: Mass gain ~ FROH",
  "Base model + sex (fixed)")

# Order by best AICc value
ms_df <- ms_df[order(ms_df$AICc), ]

# Table formatting
kable(ms_df[, c("Model", "K", "AICc", "Delta AICc")]) %>% 
  kable_styling(position = "left", full_width = FALSE)
```

```{r}
# Best model summary
summary(mod_yrborn)
```

Plotting Frequency of ROH >1Mb vs mass gain including best linear mixed effect model
```{r}
# Setting up data to incorporate predictions from best model in plot
pred_data <- data.frame(
  FROH = seq(min(mass_FlongROH$FROH),
             max(mass_FlongROH$FROH),
             length.out = 100))
pred_data$pred <- predict(mod_yrborn, newdata = pred_data, re.form = NA)

X <- model.matrix(~ FROH, pred_data)
beta <- fixef(mod_yrborn)
vcov_mat <- vcov(mod_yrborn)

pred_data$fit <- X %*% beta
pred_data$se  <- sqrt(diag(X %*% vcov_mat %*% t(X)))

#95% confidence interval to account for year born random effect in error bars 
pred_data$upper <- pred_data$fit + 1.96 * pred_data$se
pred_data$lower <- pred_data$fit - 1.96 * pred_data$se

# Making scatterplot with linear mixed effects model regression overlayed with error bars
massFROH_plot <- ggplot(mass_FlongROH, aes(x = FROH, y = massgain)) +
  geom_point() +
  geom_line(data = pred_data, aes(x = FROH, y = pred), size = 1.2) +
  geom_ribbon(data = pred_data,
              aes(y = fit, ymin = lower, ymax = upper),
              alpha = 0.2) +
  labs(x = "FROH",
       y = "Mass gain over warm season (g)") +
  eeb_theme()
massFROH_plot
```
```{r}
# ggsave("massFROH_plot.png", plot = massFROH_plot, width = 7, height = 4, units = "in", dpi=500, bg="transparent")
```

Mean ROH length vs seasonal mass gain
AIC analysis to identify best model
```{r}
# Linear model of mass gain with just FROH as fixed effect
base_mod_length <- lm(massgain ~ mean_length_kb, data = mass_FlongROH)

# Linear model of mass gain with FROH and sex as fixed effects
mod_sex_length <- lm(massgain ~ mean_length_kb + sex, data = mass_FlongROH)

# Linear mixed effects model of mass gain with FROH as fixed effect and year born as random effect
mod_yrborn_length <- lmer(massgain ~ mean_length_kb + (1|yrborn), 
                    data = mass_FlongROH, REML = FALSE)

# Linear mixed effects model of mass gain with FROH and sex as fixed effects and year born as random effect
mod_sexyrborn_length <- lmer(massgain ~ mean_length_kb + sex + (1|yrborn), 
                       data = mass_FlongROH, REML = FALSE)


# Conduct AIC analyses
ms_df_length <- as.data.frame(model.sel(base_mod_length, mod_sex_length, 
                                        mod_yrborn_length, mod_sexyrborn_length))

# Keep only useful columns
ms_df_length <- ms_df_length[, c("df", "logLik", "AICc", "delta", "weight")]

ms_df_length$AICc <- round(ms_df_length$AICc, 1)
ms_df_length$delta <- round(ms_df_length$delta, 3)

# Rename columns and models
colnames(ms_df_length) <- c("K", "logLik", "AICc", "Delta AICc", "Weight")

ms_df_length$Model <- c("Base model + birth year (random)",
  "Base model + sex (fixed) + birth year (random)",
  "Base model: Mass gain ~ FROH",
  "Base model + sex (fixed)")

# Order by best AICc value
ms_df_length <- ms_df_length[order(ms_df_length$AICc), ]

# Table formatting
kable(ms_df_length[, c("Model", "K", "AICc", "Delta AICc")]) %>% 
  kable_styling(position = "left", full_width = FALSE)
```

```{r}
# Best model summary
summary(mod_yrborn_length)
```

Plotting mean ROH length vs mass gain including best linear mixed effect model
```{r}
# Setting up data to incorporate predictions from best model in plot
pred_data_length <- data.frame(
  mean_length_kb = seq(min(mass_FlongROH$mean_length_kb),
             max(mass_FlongROH$mean_length_kb),
             length.out = 100))
pred_data_length$pred <- predict(mod_yrborn_length, newdata = pred_data_length, re.form = NA)

X <- model.matrix(~ mean_length_kb, pred_data_length)
beta <- fixef(mod_yrborn_length)
vcov_mat <- vcov(mod_yrborn_length)

pred_data_length$fit <- X %*% beta
pred_data_length$se  <- sqrt(diag(X %*% vcov_mat %*% t(X)))

#95% confidence interval to account for year born random effect in error bars 
pred_data_length$upper <- pred_data_length$fit + 1.96 * pred_data_length$se
pred_data_length$lower <- pred_data_length$fit - 1.96 * pred_data_length$se

# Making scatterplot with linear mixed effects model regression overlayed with error bars
mass_length_plot <- ggplot(mass_FlongROH, aes(x = mean_length_kb, y = massgain)) +
  geom_point() +
  geom_line(data = pred_data_length, aes(x = mean_length_kb, y = pred), size = 1.2) +
  geom_ribbon(data = pred_data_length,
              aes(y = fit, ymin = lower, ymax = upper),
              alpha = 0.2) +
  labs(x = "Mean ROH length (kb)",
       y = "Mass gain over warm season (g)") +
  eeb_theme() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), 
            axis.ticks.y = element_blank())
mass_length_plot
```

Number of ROH segments vs mass gain
AIC analysis to identify best model
```{r}
# Linear model of mass gain with just FROH as fixed effect
base_mod_num <- lm(massgain ~ NSEG, data = mass_FlongROH)

# Linear model of mass gain with FROH and sex as fixed effects
mod_sex_num <- lm(massgain ~ NSEG + sex, data = mass_FlongROH)

# Linear mixed effects model of mass gain with FROH as fixed effect and year born as random effect
mod_yrborn_num <- lmer(massgain ~ NSEG + (1|yrborn), 
                    data = mass_FlongROH, REML = FALSE)

# Linear mixed effects model of mass gain with FROH and sex as fixed effects and year born as random effect
mod_sexyrborn_num <- lmer(massgain ~ NSEG + sex + (1|yrborn), 
                       data = mass_FlongROH, REML = FALSE)


# Conduct AIC analyses
ms_df_num <- as.data.frame(model.sel(base_mod_num, mod_sex_num, 
                                        mod_yrborn_num, mod_sexyrborn_num))

# Keep only useful columns
ms_df_num <- ms_df_num[, c("df", "logLik", "AICc", "delta", "weight")]

ms_df_num$AICc <- round(ms_df_num$AICc, 1)
ms_df_num$delta <- round(ms_df_num$delta, 3)

# Rename columns and models
colnames(ms_df_num) <- c("K", "logLik", "AICc", "Delta AICc", "Weight")

ms_df_num$Model <- c("Base model + birth year (random)",
  "Base model + sex (fixed) + birth year (random)",
  "Base model + sex (fixed)",
  "Base model: Mass gain ~ FROH")

# Order by best AICc value
ms_df_num <- ms_df_num[order(ms_df_num$AICc), ]

# Table formatting
kable(ms_df_num[, c("Model", "K", "AICc", "Delta AICc")]) %>% 
  kable_styling(position = "left", full_width = FALSE)
```

```{r}
# Best model summary
summary(mod_yrborn_num)
```

Plotting number of ROH segments vs mass gain including best linear mixed effect model
```{r}
# Setting up data to incorporate predictions from best model in plot
pred_data_num <- data.frame(
  NSEG = seq(min(mass_FlongROH$NSEG),
             max(mass_FlongROH$NSEG),
             length.out = 100))
pred_data_num$pred <- predict(mod_yrborn_num, newdata = pred_data_num, re.form = NA)

X <- model.matrix(~ NSEG, pred_data_num)
beta <- fixef(mod_yrborn_num)
vcov_mat <- vcov(mod_yrborn_num)

pred_data_num$fit <- X %*% beta
pred_data_num$se  <- sqrt(diag(X %*% vcov_mat %*% t(X)))

#95% confidence interval to account for year born random effect in error bars 
pred_data_num$upper <- pred_data_num$fit + 1.96 * pred_data_num$se
pred_data_num$lower <- pred_data_num$fit - 1.96 * pred_data_num$se

# Making scatterplot with linear mixed effects model regression overlayed with error bars
mass_num_plot <- ggplot(mass_FlongROH, aes(x = NSEG, y = massgain)) +
  geom_point() +
  geom_line(data = pred_data_num, aes(x = NSEG, y = pred), size = 1.2) +
  geom_ribbon(data = pred_data_num,
              aes(y = fit, ymin = lower, ymax = upper),
              alpha = 0.2) +
  labs(x = "Number of ROH segments",
       y = "Mass gain over warm season (g)") +
  eeb_theme() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), 
            axis.ticks.y = element_blank())
  
mass_num_plot
```

```{r}
allplots <- plot_grid(massFROH_plot, mass_length_plot, mass_num_plot, ncol=3,
                      align = "h", axis = "l")
allplots
```


